---
title: "CR Cancer Proteomics"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: lumen
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(plotly)
library(DT)
library(dplyr)
library(readxl)
library(ggplot2)
library(ggrepel)
library(pdftools)

# --- Establish directories -----------
base_dir <- normalizePath(file.path(getwd()), mustWork = TRUE)
muscle_dir <- file.path(base_dir, "Muscle")
plasma_dir <- file.path(base_dir, "Plasma")

shiny::addResourcePath("muscle_gsea",
  file.path(muscle_dir, "05_analysis", "b_results"))
shiny::addResourcePath("plasma_gsea",
  file.path(plasma_dir, "05_analysis", "b_results"))
# --- Load data -----------
# Raw Data
raw_list <- list(
  Muscle = read.csv(file.path(muscle_dir, 
                              "02_normalization", 
                              "b_reports",
                              "00_raw_sample_summary_muscle.csv")),
  Plasma = read.csv(file.path(plasma_dir, 
                              "02_normalization", 
                              "b_reports",
                              "00_raw_sample_summary.csv"))
)

# Filtering Data
filter_list <- list(
  `CRE v PLA Muscle`  = read.csv(file.path(muscle_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "02_filtering_effects_CRE_muscle.csv")),
  `CRE v PLA Plasma`  = read.csv(file.path(plasma_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "02_filtering_effects_CRE.csv")),
  `SURV v CTL Muscle` = read.csv(file.path(muscle_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "02_filtering_effects_SURV_muscle.csv")),
  `SURV v CTL Plasma` = read.csv(file.path(plasma_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "02_filtering_effects_SURV.csv"))
)

# Summary Data
summary_list <- list(
  `Muscle CRE v PLA Samples` = read.csv(file.path(muscle_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "12_sample_report_CRE_muscle.csv")),
  `Muscle CRE v PLA Proteins` = read.csv(file.path(muscle_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "11_protein_report_CRE_muscle.csv")),
  `Muscle SURV v CTL Samples` = read.csv(file.path(muscle_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "12_sample_report_SURV_muscle.csv")),
  `Muscle SURV v CTL Proteins` = read.csv(file.path(muscle_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "11_protein_report_SURV_muscle.csv")),
  `Plasma CRE v PLA Samples` = read.csv(file.path(plasma_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "11_sample_report_CRE.csv")),
  `Plasma CRE v PLA Proteins` = read.csv(file.path(plasma_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "10_protein_report_CRE.csv")),
  `Plasma SURV v CTL Samples` = read.csv(file.path(plasma_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "11_sample_report_SURV.csv")),
  `Plasma SURV v CTL Proteins` = read.csv(file.path(plasma_dir, 
                                           "02_normalization", 
                                           "b_reports",
                                           "10_protein_report_SURV.csv"))
)

# Limma data
# ---- Helper to load all sheets from one file ----
load_results <- function(filepath, sheet_names) {
  setNames(
    lapply(seq_along(sheet_names), function(i) {
      read_excel(filepath, sheet = i)
    }),
    sheet_names
  )
}

muscle_sheets <- c(
  "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
  "Interaction_supp", "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV",
  "Baseline_FvM", "Training_FvM", "Training_F", "Training_M", "Interaction_sex"
)

plasma_sheets <- c(
  "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
  "Interaction_supp", "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV",
  "Baseline_FvM", "Training_FvM", "Training_F", "Training_M", "Interaction_sex"
)

results_list <- list(
  Muscle = load_results(
    file.path(muscle_dir, "05_analysis", "b_results", "limma_results_all_muscle.xlsx"),
    muscle_sheets
  ),
  Plasma = load_results(
    file.path(plasma_dir, "05_analysis", "b_results", "limma_results_all_plasma.xlsx"),
    plasma_sheets
  )
)

# PCA Data
pca_list <- list(
  Muscle_SURV = readRDS(file.path(muscle_dir, "03_imputation", "c_data", "01_imputed_DAList_SURV.RDS")),
  Muscle_CRE  = readRDS(file.path(muscle_dir, "03_imputation", "c_data", "01_imputed_DAList_CRE.RDS")),
  Plasma_SURV = readRDS(file.path(plasma_dir, "03_imputation", "c_data", "01_imputed_DAList_SURV.RDS")),
  Plasma_CRE  = readRDS(file.path(plasma_dir, "03_imputation", "c_data", "01_imputed_DAList_CRE.RDS"))
)

# Normalized Data (for regressions)


  
```

```{css, echo=FALSE}
/* Flexdashboard sidebar fix across all tabs */
.sidebar.section.sidebar {
  top: 51px !important;
}

/* Fallback — targets the actual rendered sidebar div */
div[class*="sidebar"] {
  margin-top: 0 !important;
  padding-top: 10px !important;
}

.navbar ~ .container-fluid .section.sidebar {
  top: 51px !important;
  position: fixed !important;
}
```


# Raw Data Processing {data-navmenu="Data" data-icon="fa-cogs"}

## Sidebar {.sidebar}
```{r sidebar1}
# --- Sample selector ---
selectInput("sample_1",
  label = "Sample",
  choices = c("Muscle", "Plasma"),
  selected = "Muscle")

# --- Dataset selector ---
selectInput("dataset_1",
  label = "Dataset",
  choices = c("CRE v PLA", "SURV v CTL"),
  selected = "CRE v PLA")
```
## Column {data-width=500}
### Raw Data Summary
``` {r valuebox1}
valueBoxOutput("raw_summary_box")
```
### Raw Data - Sample Summary
```{r raw data table}
DTOutput("raw_table")
```

## Column {data-width=500}
### Filtering Plot
```{r filter plot}
plotOutput("filter_plot")
```
### Filtering
```{r filter table}
DTOutput("filter_table")
```

```{r server1, context="server"}
# Raw Data Summary Table
make_raw_sample_table <- function(df) {
  df %>%
    dplyr::select(-1) %>%
    dplyr::select(-cancer_time, -supp_time)
}

output$raw_table <- renderDT({
  key <- paste(input$sample_1)
  df <- raw_list[[key]]
  req(df)
  
  make_raw_sample_table(df) %>%
     datatable(
      options  = list(dom = "t", pageLength = 10, ordering = FALSE),
      rownames = FALSE
    )
})

# Filter Table 
output$filter_table <- renderDT({
  key <- paste(input$dataset_1, input$sample_1)
  df <- filter_list[[key]]
  req(df)
  
  df %>% datatable(
    options  = list(dom = "t", pageLength = 10, ordering = FALSE),
      rownames = FALSE
  )
})

# Filter Bar Plot
make_filter_plot <- function(df, label) {
  ggplot(mutate(df, step = factor(step, levels = step)), aes(x = step, y = n_after)) +
    geom_col(fill = "steelblue", width = 0.5) +
    geom_text(aes(label = n_after), vjust = -0.3, size = 3) +
    labs(x = NULL, y = "Number of Proteins",
         title = paste("Proteins Retained Through Filtering -", label)) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
}

output$filter_plot <- renderPlot({
  key <- paste(input$dataset_1, input$sample_1)
  df <- filter_list[[key]]
  req(df)
  make_filter_plot(df, input$dataset_1)
})

# Raw Data Summary Value Box
output$raw_summary_box <- renderValueBox({
  key_proteins <- paste(input$dataset_1, input$sample_1)
  key_samples <- paste(input$sample_1)
  
  n_proteins <- filter_list[[key_proteins]] %>%
    dplyr::slice_head(n = 1) %>%
    pull(n_after)
  
  dfr <- raw_list[[key_samples]]
  
  n_samples <- sum(dfr$n)
  
  valueBox(
    value = paste0(n_proteins, " proteins x ", n_samples, " samples"),
    color = "primary"
  )
})
```


# Normalized Data Summary {data-navmenu="Data" data-icon="fa-table"}

## Sidebar {.sidebar}
```{r sidebar2}
# --- Sample selector ---
selectInput("sample_2",
  label = "Sample",
  choices = c("Muscle", "Plasma"),
  selected = "Muscle")

# --- Dataset selector ---
selectInput("dataset_2",
  label = "Dataset",
  choices = c("CRE v PLA", "SURV v CTL"),
  selected = "CRE v PLA")

# --- Summary selector ---
selectInput("summary",
  label = "Summary",
  choices = c("Samples", "Proteins"),
  selected = "Samples")
```

## Column {data-width=650}
### Data Summary
```{r valuebox2}
valueBoxOutput("data_value_box")
```
### Data Summary
```{r table1}
DTOutput("sum_table")
```

```{r server2, context="server"}
# ---- Summary data table ----
make_sample_summary_cre <- function(df) {
  # Pull values
  total    <- df$total_samples
  n_cre_t1 <- df$n_CRE_T1;  n_cre_t2 <- df$n_CRE_T2
  n_pla_t1 <- df$n_PLA_T1;  n_pla_t2 <- df$n_PLA_T2
  n_paired <- df$n_paired * 2
  n_pairs  <- df$n_paired 
  n_unpaired <- df$n_unpaired
  
  # Paired T1/T2 per group = n_paired_CRE / n_paired_PLA each split evenly
  p_cre_t1 <- df$n_paired_CRE;  p_cre_t2 <- df$n_paired_CRE
  p_pla_t1 <- df$n_paired_PLA;  p_pla_t2 <- df$n_paired_PLA
  
  # Unpaired = All - Paired per cell
  u_cre_t1 <- n_cre_t1 - p_cre_t1;  u_cre_t2 <- n_cre_t2 - p_cre_t2
  u_pla_t1 <- n_pla_t1 - p_pla_t1;  u_pla_t2 <- n_pla_t2 - p_pla_t2
  
  tibble::tibble(
    ` `      = c("All", paste0("Paired (", n_pairs, " pairs)"), "Unpaired"),
    `n`      = c(total,   n_paired,   n_unpaired),
    `CRE T1` = c(n_cre_t1, p_cre_t1, u_cre_t1),
    `CRE T2` = c(n_cre_t2, p_cre_t2, u_cre_t2),
    `PLA T1` = c(n_pla_t1, p_pla_t1, u_pla_t1),
    `PLA T2` = c(n_pla_t2, p_pla_t2, u_pla_t2)
  )
}

# Summary Table Function 
make_sample_summary_surv <- function(df) {
  # Pull values
  total     <- df$total_samples
  n_surv_t1 <- df$n_CRE_T1 + df$n_PLA_T1
  n_surv_t2 <- df$n_CRE_T2 + df$n_PLA_T2
  n_ctl_t1  <- df$n_CTL   
  n_paired  <- df$n_paired * 2
  n_pairs   <- df$n_paired 
  n_unpaired <- df$n_unpaired
  
  # Paired T1/T2 per group = n_paired_CRE / n_paired_PLA each split evenly
  p_surv_t1 <- df$n_paired  
  p_surv_t2 <- df$n_paired
  p_ctl_t1  <- NA 
  
  # Unpaired = All - Paired per cell
  u_surv_t1 <- n_surv_t1 - p_surv_t1
  u_surv_t2 <- n_surv_t2 - p_surv_t2
  u_ctl_t1 <- n_ctl_t1 
  
  tibble::tibble(
    ` `       = c("All", paste0("Paired (", n_pairs, " pairs)"), "Unpaired"),
    `n`       = c(total,   n_paired,   n_unpaired),
    `SURV T1` = c(n_surv_t1, p_surv_t1, u_surv_t1),
    `SURV T2` = c(n_surv_t2, p_surv_t2, u_surv_t2),
    `CTL T1`  = c(n_ctl_t1, p_ctl_t1, u_ctl_t1)
  )
}

make_protein_summary_cre <- function(df) {
  tibble::tibble(
    ` `          = c("Total Proteins", "Complete Proteins", "Incomplete Proteins"),
    `n`          = c(df$Total, df$Complete, df$Incomplete),
    `% of Total` = c(paste0(round(df$Total / df$Total * 100, 1), "%"),
                     paste0(round(df$Complete / df$Total * 100, 1), "%"),
                     paste0(round(df$Incomplete / df$Total * 100, 1), "%"))
  )
}

make_protein_summary_surv <- function(df) {
  tibble::tibble(
    ` `          = c("Total Proteins", "Complete Proteins", "Incomplete Proteins"),
    `n`          = c(df$Total, df$Complete, df$Incomplete),
    `% of Total` = c(paste0(round(df$Total / df$Total * 100, 1), "%"),
                     paste0(round(df$Complete / df$Total * 100, 1), "%"),
                     paste0(round(df$Incomplete / df$Total * 100, 1), "%"))
  )
}

# Lookup table
summary_builders <- list(
  `Muscle CRE v PLA Samples`   = make_sample_summary_cre,
  `Muscle CRE v PLA Proteins`  = make_protein_summary_cre,
  `Muscle SURV v CTL Samples`  = make_sample_summary_surv,
  `Muscle SURV v CTL Proteins` = make_protein_summary_surv,
  `Plasma CRE v PLA Samples`   = make_sample_summary_cre,
  `Plasma CRE v PLA Proteins`  = make_protein_summary_cre,
  `Plasma SURV v CTL Samples`  = make_sample_summary_surv,
  `Plasma SURV v CTL Proteins` = make_protein_summary_surv
)

# Summary Table Output
output$sum_table <- renderDT({
  key <- paste(input$sample_2, input$dataset_2, input$summary)
  df <- summary_list[[key]] %>% dplyr::select(-1)
  builder <- summary_builders[[key]]
  req(df, builder)
  
  builder(df) %>%
    datatable(
      options  = list(dom = "t", pageLength = 10, ordering = FALSE),
      rownames = FALSE
    )
})

# Normalized Data Summary Box
output$data_value_box <- renderValueBox({
  key_pro <- paste(input$dataset_2, input$sample_2)
  key_sam <- paste(input$sample_2, input$dataset_2, "Samples")
  
  n_pro <- filter_list[[key_pro]] %>%
    dplyr::slice_tail(n = 1) %>%
    pull(n_after)
  
  n_sam <- summary_list[[key_sam]] %>%
    dplyr::select(-1) %>%
    pull(total_samples)
  
   valueBox(
    value = paste0(n_pro, " proteins x ", n_sam, " samples"),
    color = "primary"
  )
})

# ---- Comparison choices lookup ----
comparison_choices <- list(
  `Muscle CRE v PLA` = c(
    "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
    "Interaction_supp", "Baseline_FvM", "Training_FvM", "Training_F",
    "Training_M", "Interaction_sex"
  ),
  `Muscle SURV v CTL` = c(
    "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV"
  ),
  `Plasma CRE v PLA` = c(
    "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
    "Interaction_supp", "Baseline_FvM", "Training_FvM", "Training_F",
    "Training_M", "Interaction_sex"
  ),
  `Plasma SURV v CTL` = c(
    "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV"
  )
)

# Comparison Selector UI
output$comparison_ui <- renderUI({
  key     <- paste(input$sample_3, input$dataset_3)
  choices <- comparison_choices[[key]]
  req(choices)
  selectInput("comparison",
    label    = "Comparison",
    choices  = choices,
    selected = choices[1])
})
```

# Volcano Plots {data-navmenu="Outputs" data-icon="fa-volcano"}
## Sidebar {.sidebar}
```{r sidebar3}
# --- Sample selector ---
selectInput("sample_3",
  label = "Sample",
  choices = c("Muscle", "Plasma"),
  selected = "Muscle")

# --- Dataset selector ---
selectInput("dataset_3",
  label = "Dataset",
  choices = c("CRE v PLA", "SURV v CTL"),
  selected = "CRE v PLA")

# --- Comparison selector ---
uiOutput("comparison_ui")

hr()

# --- p-value type ---
radioButtons("pval_type",
  label    = "P-value type",
  choices  = c("Raw p-value" = "P.Value", 
               "Adjusted p-value" = "adj.P.Val",
               "Adjusted Pi-value (Xiao)" = "adj.Pi.Val"),
  selected = "adj.P.Val"
)

# --- Thresholds ---
sliderInput("pval_thresh",
  label = "P-value threshold",
  min = 0.01, max = 0.1, value = 0.05, step = 0.01
)

sliderInput("logfc_thresh",
  label = "|logFC| threshold",
  min = 0, max = 1, value = 0.6, step = 0.1
)

hr()

# --- Optional: colour by significance ---
checkboxInput("colour_sig", "Colour by significance", value = TRUE)
```

## Column {data-width=650}
### Volcano Plot
```{r volcano}
plotlyOutput("volcano_plot")
```

## Column {data-width=350}
### Significant Proteins
```{r table}
DTOutput("sig_table")
```

### Summary
```{r summary}
verbatimTextOutput("summary_text")
```

```{r server3, context="server"}

# ---- Reactive: filtered dataset ----
dat <- reactive({
  req(input$comparison, input$sample_3)
  df <- results_list[[input$sample_3]][[input$comparison]]
  req(df)
  df$pval_used <- df[[input$pval_type]]
  df$sig <- df$pval_used < input$pval_thresh & abs(df$logFC) > input$logfc_thresh
  df$direction <- ifelse(!df$sig, "ns",
                    ifelse(df$logFC > 0, "up", "down"))
  df
})

# ---- Volcano plot ----
output$volcano_plot <- renderPlotly({
  df <- dat()
  
  colour_map <- c("up" = "#e63946", "down" = "turquoise3", "ns" = "#adb5bd")
  
  p <- plot_ly(df,
    x = ~logFC,
    y = ~-log10(pval_used),
    type = "scatter", mode = "markers",
    color = if (input$colour_sig) ~direction else I("#adb5bd"),
    colors = colour_map,
    text = ~paste0("<b>", gene, "</b><br>",
                   description, "<br>",
                   "logFC: ", round(logFC, 3), "<br>",
                   "p: ", signif(pval_used, 3)),
    hoverinfo = "text",
    marker = list(size = 6, opacity = 0.75)
  ) %>%
    layout(
      title  = input$comparison,
      xaxis  = list(title = "log2 Fold Change",
                    zeroline = TRUE),
      yaxis  = list(title = paste0("-log10(", input$pval_type, ")")),
      shapes = list(
        # vertical logFC lines
        list(type="line", x0=input$logfc_thresh,  x1=input$logfc_thresh,
             y0=0, y1=1, yref="paper",
             line=list(dash="dot", color="grey")),
        list(type="line", x0=-input$logfc_thresh, x1=-input$logfc_thresh,
             y0=0, y1=1, yref="paper",
             line=list(dash="dot", color="grey")),
        # horizontal p-value line
        list(type="line", x0=0, x1=1, xref="paper",
             y0=-log10(input$pval_thresh), y1=-log10(input$pval_thresh),
             line=list(dash="dot", color="grey"))
      ),
      legend = list(title = list(text = "Direction"))
    )
  p
})

# ---- Significant proteins table ----
output$sig_table <- renderDT({
  dat() %>%
    filter(sig) %>%
    select(description, gene, logFC, pval_used, direction) %>%
    arrange(pval_used) %>%
    rename("Protein" = description, "Gene" = gene, "FC" = logFC,
           "p-value" = pval_used, "Direction" = direction) %>%
    mutate(across(where(is.numeric), ~signif(.x, 3))) %>%
    datatable(options = list(pageLength = 10, dom = "tp"),
              rownames = FALSE)
}, server = FALSE)

# ---- Summary text ----
output$summary_text <- renderText({
  df  <- dat()
  up  <- sum(df$direction == "up",   na.rm = TRUE)
  dn  <- sum(df$direction == "down", na.rm = TRUE)
  tot <- nrow(df)
  paste0(
    "Comparison: ",    input$comparison, "\n",
    "Total proteins: ", tot, "\n",
    "Up-regulated: ",   up,  "\n",
    "Down-regulated: ", dn,  "\n",
    "logFC cutoff: ±",  input$logfc_thresh, "\n",
    input$pval_type, " < ", input$pval_thresh
  )
})
```

# Pathway Analysis {data-navmenu="Outputs" data-icon="fa-chart-bar"}
## Sidebar {.sidebar}
```{r sidebar4}
# --- Sample selector ---
selectInput("sample_4",
  label = "Sample",
  choices = c("Muscle", "Plasma"),
  selected = "Muscle")

# --- Dataset selector ---
selectInput("dataset_4",
  label = "Dataset",
  choices = c("CRE v PLA", "SURV v CTL"),
  selected = "SURV v CTL")

# --- Comparison selector ---
uiOutput("comparison_ui2")

# --- Pathway selectors ---
selectInput("analysis",
             label = "Analysis",
             choices = c("GSEA", "ORA"),
             selected = "GSEA")

uiOutput("ont_ui")
```

## Column {width = 1000}
### Plot:
```{r gsea}
imageOutput("gsea_plot")
```

```{r server4, context="server"}
# ---- Comparison choices lookup ----
comparison_choices <- list(
  `Muscle CRE v PLA` = c(
    "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
    "Interaction_supp", "Baseline_FvM", "Training_FvM", "Training_F",
    "Training_M", "Interaction_sex"
  ),
  `Muscle SURV v CTL` = c(
    "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV"
  ),
  `Plasma CRE v PLA` = c(
    "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
    "Interaction_supp", "Baseline_FvM", "Training_FvM", "Training_F",
    "Training_M", "Interaction_sex"
  ),
  `Plasma SURV v CTL` = c(
    "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV"
  )
)

output$comparison_ui2 <- renderUI({
  key     <- paste(input$sample_4, input$dataset_4)
  choices <- comparison_choices[[key]]
  req(choices)
  selectInput("comparison_2",
    label    = "Comparison",
    choices  = choices,
    selected = choices[1])
})

ont_choices <- list(
  GSEA = c("BP", "CC", "MF", "hallmark", "KEGG", "KEGG_ridge"),
  ORA = c("BP", "CC", "MF")
)

output$ont_ui <- renderUI({
  key <- input$analysis
  choices <- ont_choices[[key]]
  req(choices)
  selectInput("ont",
              label = "Gene Ontology",
              choices = choices,
              selected = choices[1])
})

output$gsea_plot <- renderImage({
  req(input$sample_4, input$dataset_4, input$comparison_2, input$analysis, input$ont)
  
  gsea_dir <- if (input$sample_4 == "Muscle") muscle_dir else plasma_dir
  comp_dir <- if (input$dataset_4 == "CRE v PLA") "CREvPLA" else "SURVvCTL"
  go_dir   <- if (input$analysis == "GSEA") "GSEA" else "ORA"
  fname    <- if (input$analysis =="GSEA") 
    sprintf("GSEA_%s_%s.pdf", input$ont, input$comparison_2) else
      sprintf("ORA_GO_%s_bar_%s.pdf", input$ont, input$comparison_2)
  pdf_path <- file.path(gsea_dir, "05_analysis", "b_results", comp_dir, go_dir, fname)
  
  req(file.exists(pdf_path))
  
  # Convert first page of PDF to a temp PNG
  png_path <- tempfile(fileext = ".png")
  pdftools::pdf_convert(pdf_path, format = "png", pages = 1,
                        filenames = png_path, dpi = 150)
  
 img_info <- pdftools::pdf_pagesize(pdf_path)
 aspect   <- img_info$height[1] / img_info$width[1]

 list(
   src         = png_path,
   contentType = "image/png",
   width       = "100%",
   height      = paste0(round(800 * aspect), "px")  # scales to actual proportions
)
}, deleteFile = TRUE)   # TRUE so temp PNGs are cleaned up automatically
```

# Scatter Plots {data-navmenu="Outputs" data-icon="fa-border-none"}
## Sidebar {.sidebar}
```{r}
selectInput("sample_5",
            label = "Sample",
            choices = c("Muscle", "Plasma"),
            selected = "Muscle")

uiOutput("concordance_ui")
```

## Column {width = 1000}
### Scatter Plot
```{r}
plotlyOutput("scatter")
```

```{r server5, context="server"}
# comparison choices
concordance_choices <- list(
  Muscle = c(
    "Sex (F vs M)"            = "FvM",
    "Supplement (CRE vs PLA)" = "CREvPLA",
    "Survivors (T1 vs T2)"    = "SURVvCTL"
  ),
  Plasma = c(
    "Sex (F vs M)"            = "FvM",
    "Supplement (CRE vs PLA)" = "CREvPLA",
    "Survivors (T1 vs T2)"    = "SURVvCTL"
  )
)

output$concordance_ui <- renderUI({
  selectInput("concordance_comparison", "Comparison:",
    choices = concordance_choices[[input$sample_5]])
})

# Prepare Data Function
prep_cord <- function(df, label) {
  df %>%
    dplyr::select(gene, logFC, adj.P.Val) %>%
    dplyr::rename(
      !!paste0("logFC_",   label) := logFC,
      !!paste0("adj.P.Val_", label) := adj.P.Val
    ) %>%
    mutate(!!paste0("signif_", label) := .data[[paste0("adj.P.Val_", label)]] < 0.05)
}

# Pre-prepare all datasets once
# Reactive cord_data based on sample
cord_data <- reactive({
  req(input$sample_5)
  src <- results_list[[input$sample_5]]
  
  if (input$sample_5 == "Muscle") {
    list(
      F       = prep_cord(src$Training_F,          "F"),
      M       = prep_cord(src$Training_M,          "M"),
      CRE     = prep_cord(src$Training_CRE,        "CRE"),
      PLA     = prep_cord(src$Training_PLA,        "PLA"),
      SURV_T1 = prep_cord(src$Baseline_SURVvCTL,   "SURV_T1"),
      SURV_T2 = prep_cord(src$Training_SURVvCTL,   "SURV_T2")
    )
  } else {
    list(
      F       = prep_cord(src$Training_F,          "F"),
      M       = prep_cord(src$Training_M,          "M"),
      CRE     = prep_cord(src$Training_CRE,        "CRE"),
      PLA     = prep_cord(src$Training_PLA,        "PLA"),
      SURV_T1 = prep_cord(src$Baseline_SURVvCTL,   "SURV_T1"),
      SURV_T2 = prep_cord(src$Training_SURVvCTL,   "SURV_T2")
    )
  }
})


# Build Plots
plot_concordance <- function(df1, df2, label1, label2, title) {
  
  col1 <- paste0("logFC_", label1)
  col2 <- paste0("logFC_", label2)
  sig1 <- paste0("signif_", label1)
  sig2 <- paste0("signif_", label2)
  
  cord <- inner_join(df1, df2, by = "gene") %>%
    mutate(
      sig_group = case_when(
        .data[[sig1]] & .data[[sig2]]  ~ "Both significant",
        .data[[sig1]] & !.data[[sig2]] ~ paste0(label1, " only"),
        !.data[[sig1]] & .data[[sig2]] ~ paste0(label2, " only"),
        TRUE                           ~ "Neither"
      )
    )

  groups <- list(
    list(name = "Both significant",        color = "purple",  opacity = 1),
    list(name = paste0(label1, " only"),   color = "red",     opacity = 1),
    list(name = paste0(label2, " only"),   color = "blue",    opacity = 1),
    list(name = "Neither",                 color = "grey2", opacity = 0.7)
  )

  x_range <- range(cord[[col1]], na.rm = TRUE)
  y_range <- range(cord[[col2]], na.rm = TRUE)
  pad     <- 0.1
  xmin <- x_range[1] - pad; xmax <- x_range[2] + pad
  ymin <- y_range[1] - pad; ymax <- y_range[2] + pad

  p <- plot_ly() %>%
    
    # --- Quadrant shading ---
    add_trace(type = "scatter", mode = "none", inherit = FALSE,
          fill = "toself",
          fillcolor = "rgba(999,999,999,0.1)",
          x = c(0, xmax, xmax, 0), y = c(0, 0, ymax, ymax),
          showlegend = FALSE, hoverinfo = "none") %>%
    add_trace(type = "scatter", mode = "none", inherit = FALSE,
          fill = "toself",
          fillcolor = "rgba(999,999,999,0.1)",
          x = c(xmin, 0, 0, xmin), y = c(ymin, ymin, 0, 0),
          showlegend = FALSE, hoverinfo = "none") %>%
    add_trace(type = "scatter", mode = "none", inherit = FALSE,
          fill = "toself",
          fillcolor = "rgba(0,0,0,0.1)",
          x = c(0, xmax, xmax, 0), y = c(ymin, ymin, 0, 0),
          showlegend = FALSE, hoverinfo = "none") %>%
    add_trace(type = "scatter", mode = "none", inherit = FALSE,
          fill = "toself",
          fillcolor = "rgba(0,0,0,0.1)",
          x = c(xmin, 0, 0, xmin), y = c(0, 0, ymax, ymax),
          showlegend = FALSE, hoverinfo = "none") %>%

    # --- Quadrant labels ---
    add_annotations(x = xmax, y = ymax, text = paste("UP", label1, "&", label2),
                    xanchor = "right", yanchor = "top",
                    font = list(color = "darkgreen", size = 11), showarrow = FALSE) %>%
    add_annotations(x = xmin, y = ymin, text = paste("DOWN", label1, "&", label2),
                    xanchor = "left", yanchor = "bottom",
                    font = list(color = "darkgreen", size = 11), showarrow = FALSE) %>%
    add_annotations(x = xmax, y = ymin, text = paste0("UP ", label1, " / DOWN ", label2),
                    xanchor = "right", yanchor = "bottom",
                    font = list(color = "steelblue", size = 11), showarrow = FALSE) %>%
    add_annotations(x = xmin, y = ymax, text = paste0("UP ", label2, " / DOWN ", label1),
                    xanchor = "left", yanchor = "top",
                    font = list(color = "red", size = 11), showarrow = FALSE)

  # --- One trace per group ---
  for (g in groups) {
    d <- cord[cord$sig_group == g$name, ]
    p <- add_trace(p,
      data      = d,
      type      = "scatter", mode = "markers",
      x         = ~.data[[col1]],
      y         = ~.data[[col2]],
      name      = g$name,
      marker    = list(color = g$color, size = 7, opacity = g$opacity),
      text      = ~paste0("<b>", gene, "</b><br>",
                          label1, " logFC: ", round(.data[[col1]], 3), "<br>",
                          label2, " logFC: ", round(.data[[col2]], 3), "<br>",
                          "Signif: ", sig_group),
      hoverinfo = "text"
    )
  }

  p %>% layout(
    title     = list(text = title, font = list(size = 14)),
    xaxis     = list(title = paste(label1, "(logFC)"), zeroline = TRUE,
                     zerolinecolor = "grey", range = c(xmin, xmax)),
    yaxis     = list(title = paste(label2, "(logFC)"), zeroline = TRUE,
                     zerolinecolor = "grey", range = c(ymin, ymax)),
    legend    = list(title = list(text = "Significant DE")),
    shapes    = list(
      list(type = "line", x0 = 0, x1 = 0, y0 = ymin, y1 = ymax,
           line = list(color = "grey60", dash = "dash", width = 1)),
      list(type = "line", x0 = xmin, x1 = xmax, y0 = 0, y1 = 0,
           line = list(color = "grey60", dash = "dash", width = 1))
    ),
    hovermode = "closest"
  )
}

# Plot Output
output$scatter <- renderPlotly({
  req(input$concordance_comparison, input$sample_5)
  cd <- cord_data()
  
  switch(input$concordance_comparison,
    FvM      = plot_concordance(cd$F,       cd$M,       "F",       "M",       paste("Sex Comparison -",        input$sample_5)),
    CREvPLA  = plot_concordance(cd$CRE,     cd$PLA,     "CRE",     "PLA",     paste("Supplement Comparison -", input$sample_5)),
    SURVvCTL = plot_concordance(cd$SURV_T1, cd$SURV_T2, "SURV_T1", "SURV_T2", paste("SURV v CTL -",           input$sample_5))
  )
})
```

# PCA {data-navmenu="Outputs" data-icon="fa-circle-dot"}
## Sidebar {.sidebar}
```{r}
selectInput("sample_6",
            label = "Sample",
            choices = c("Muscle", "Plasma"),
            selected = "Muscle")

selectInput("dataset_6",
            label = "Dataset",
            choices = c("CRE v PLA", "SURV v CTL"),
            selected = "SURV v CTL")

uiOutput("pca_color_ui")
```

## Column {width = 1000}
### PCA Plot
```{r}
plotlyOutput("pca_plot")
```

```{r server6, context="server"}

# Dynamic color_by choices based on dataset
output$pca_color_ui <- renderUI({
  choices <- if (input$dataset_6 == "SURV v CTL") {
    c("Cancer & Time" = "cancer_time", "Cancer" = "cancer")
  } else {
    c("Supplement & Time" = "supp_time", 
      "Supplement" = "supp", 
      "Time" = "timepoint",
      "Sex" = "sex",
      "Sex & Time" = "sex_time")
  }
  selectInput("pca_color", label = "Colour by", choices = choices)
})

# PCA ellipse helper
compute_ellipse <- function(x, y, level = 0.95, n = 100) {
  if (length(x) < 3) return(NULL)
  tryCatch({
    fit    <- MASS::cov.trob(cbind(x, y), nu = length(x) - 1)
    center <- fit$center
    cov    <- fit$cov
    eigen  <- eigen(cov)
    chi2   <- qchisq(level, df = 2)
    angles <- seq(0, 2 * pi, length.out = n)
    circle <- cbind(cos(angles), sin(angles))
    ellipse <- t(center + t(circle %*% (sqrt(chi2) * t(chol(cov)))))
    data.frame(x = ellipse[, 1], y = ellipse[, 2])
  }, error = function(e) NULL)
}

# Hex-to-rgba Helper
grp_to_rgba <- function(col, alpha = 0.1) {
  rgb_vals <- col2rgb(col) / 255
  sprintf("rgba(%s,%s,%s,%s)",
          round(rgb_vals[1] * 255),
          round(rgb_vals[2] * 255),
          round(rgb_vals[3] * 255),
          alpha)
}

# PCA Plot Output
output$pca_plot <- renderPlotly({
  req(input$sample_6, input$dataset_6, input$pca_color)
  
  key      <- paste0(input$sample_6, "_", if (input$dataset_6 == "CRE v PLA") "CRE" else "SURV")
  dal      <- pca_list[[key]]
  mat      <- dal$data
  meta     <- dal$metadata
  color_by <- input$pca_color
  
  pca     <- prcomp(t(mat), scale. = TRUE)
  var_exp <- round(summary(pca)$importance[2, 1:2] * 100, 1)
  
  pca_df <- as.data.frame(pca$x[, 1:2]) %>%
    tibble::rownames_to_column("sample_id") %>%
    left_join(meta, by = "sample_id")
  
  groups  <- unique(pca_df[[color_by]])
  palette <- setNames(RColorBrewer::brewer.pal(max(3, length(groups)), "Set1")[seq_along(groups)], groups)

  p <- plot_ly()

  for (grp in groups) {
    d     <- pca_df[pca_df[[color_by]] == grp, ]
    col   <- palette[[grp]]
    ell   <- compute_ellipse(d$PC1, d$PC2)

    # Ellipse
    if (!is.null(ell)) {
      p <- add_trace(p,
        x          = ell$x, y = ell$y,
        type       = "scatter", mode = "lines",
        line       = list(color = col, width = 1.5, dash = "solid"),
        fill       = "toself",
        fillcolor  = paste0(grp_to_rgba(col, 0.1)),
        name       = grp,
        showlegend = FALSE,
        hoverinfo  = "skip"
      )
    }

    # Points
    p <- add_trace(p,
      x         = d$PC1, y = d$PC2,
      type      = "scatter", mode = "markers",
      name      = grp,
      marker    = list(color = col, size = 8),
      text      = paste0("<b>", d$sample_id, "</b><br>", color_by, ": ", d[[color_by]]),
      hoverinfo = "text"
    )
  }

  p %>% layout(
    title  = list(text = paste(input$sample_6, "-", input$dataset_6), font = list(size = 14)),
    xaxis  = list(title = paste0("PC1 (", var_exp[1], "%)")),
    yaxis  = list(title = paste0("PC2 (", var_exp[2], "%)")),
    legend = list(title = list(text = color_by))
  )
})
```

# DE Summary {data-navmenu="Outputs" data-icon="fa-bar-chart"}
## Sidebar {.sidebar}
```{r}
selectInput("sample_7",
            label = "Sample",
            choices = c("Muscle", "Plasma"),
            selected = "Muscle")

selectInput("dataset_7",
            label = "Dataset",
            choices = c("CRE v PLA", "SURV v CTL"),
            selected = "SURV v CTL")

uiOutput("comparison_ui7")

hr()

radioButtons("pval_type_7",
  label   = "P-value type",
  choices = c("Raw p-value"              = "P.Value",
              "Adjusted p-value"         = "adj.P.Val",
              "Adjusted π-value (Xiao)" = "adj.Pi.Val"),
  selected = "adj.P.Val"
)

sliderInput("pval_thresh_7",
  label = "P-value threshold",
  min = 0.01, max = 0.1, value = 0.05, step = 0.01
)

sliderInput("logfc_thresh_7",
  label = "|logFC| threshold",
  min = 0, max = 1, value = 0.0, step = 0.1
)
```

## Column {width = 1000}
### DE Summary
```{r}
plotlyOutput("de_bar")
```

```{r server7, context="server"}

comparison_choices_7 <- list(
  `Muscle CRE v PLA` = c(
    "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
    "Interaction_supp", "Baseline_FvM", "Training_FvM", "Training_F",
    "Training_M", "Interaction_sex"
  ),
  `Muscle SURV v CTL` = c(
    "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV"
  ),
  `Plasma CRE v PLA` = c(
    "Baseline_CREvPLA", "Training_CREvPLA", "Training_CRE", "Training_PLA",
    "Interaction_supp", "Baseline_FvM", "Training_FvM", "Training_F",
    "Training_M", "Interaction_sex"
  ),
  `Plasma SURV v CTL` = c(
    "Baseline_SURVvCTL", "Training_SURVvCTL", "Training_SURV"
  )
)

output$comparison_ui7 <- renderUI({
  key     <- paste(input$sample_7, input$dataset_7)
  choices <- comparison_choices_7[[key]]
  req(choices)
  checkboxGroupInput("comparisons_7",
    label    = "Comparisons",
    choices  = choices,
    selected = choices
  )
})

output$de_bar <- renderPlotly({
  req(input$comparisons_7, input$sample_7, input$pval_type_7)

  # --- Build summary across selected comparisons ---
  summary_df <- lapply(input$comparisons_7, function(comp) {
    df <- results_list[[input$sample_7]][[comp]]
    req(df)
    df %>%
      mutate(
        pval_used = .data[[input$pval_type_7]],
        sig       = pval_used < input$pval_thresh_7 & abs(logFC) > input$logfc_thresh_7,
        direction = ifelse(!sig, "ns", ifelse(logFC > 0, "Up", "Down"))
      ) %>%
      summarise(
        Up   =  sum(direction == "Up",   na.rm = TRUE),
        Down = -sum(direction == "Down", na.rm = TRUE)  # negative for downward bars
      ) %>%
      mutate(comparison = comp)
  }) %>%
    bind_rows() %>%
    tidyr::pivot_longer(cols = c(Up, Down), names_to = "direction", values_to = "n") %>%
    mutate(comparison = factor(comparison, levels = input$comparisons_7))

  plot_ly(summary_df,
    x         = ~comparison,
    y         = ~n,
    color     = ~direction,
    colors    = c("Up" = "#e63946", "Down" = "turquoise3"),
    type      = "bar",
    text      = ~paste0(direction, ": ", abs(n)),
    hoverinfo = "text"
  ) %>%
    layout(
      barmode  = "relative",
      title    = list(text = paste(input$sample_7, "-", input$dataset_7), font = list(size = 14)),
      xaxis    = list(title = "Comparison", tickangle = -35),
      yaxis    = list(title = "Number of Proteins",
                      tickvals = NULL,
                      ticktext = NULL),
      legend   = list(title = list(text = "Direction")),
      shapes   = list(
        list(type = "line", x0 = -0.5, x1 = length(input$comparisons_7) - 0.5,
             y0 = 0, y1 = 0,
             line = list(color = "black", width = 1))
      )
    ) %>%
    # Fix y axis labels to show absolute values
    layout(yaxis = list(
      tickmode = "auto",
      tickformat = ",.0f",
      title = "Number of Proteins"
    ))
})
```

# Venn Diagrams {data-navmenu="Outputs" data-icon="fa-circle"}
## Sidebar {.sidebar}
```{r sidebar8}
selectInput("sample_8",
            label = "Sample",
            choices = c("Muscle", "Plasma"),
            selected = "Muscle")

selectInput("dataset_8",
            label = "Dataset",
            choices = c("CRE v PLA", "SURV v CTL"),
            selected = "SURV v CTL")
```
## Column {width = 1000}
### Venn
```{r venn_output}
imageOutput("venn")
```

```{r server8, context="server"}
output$venn <- renderImage({
  req(input$sample_8, input$dataset_8)
  
  samp_dir <- if (input$sample_8 == "Muscle") muscle_dir else plasma_dir
  data_dir <- if (input$dataset_8 == "SURV v CTL") "SURVvCTL" else "CREvPLA"
  vname    <- sprintf("venn_%s_proteins.pdf", data_dir) 
  pdf_path <- file.path(samp_dir, "05_analysis", "b_results", data_dir, "plots", "venns", vname)
  
  req(file.exists(pdf_path))
  
  # Convert first page of PDF to a temp PNG
  png_path <- tempfile(fileext = ".png")
  pdftools::pdf_convert(pdf_path, format = "png", pages = 1,
                        filenames = png_path, dpi = 150)
  
 img_info <- pdftools::pdf_pagesize(pdf_path)
 aspect   <- img_info$height[1] / img_info$width[1]

 list(
   src         = png_path,
   contentType = "image/png",
   width       = "100%",
   height      = paste0(round(800 * aspect), "px")  # scales to actual proportions
)
}, deleteFile = TRUE)   # TRUE so temp PNGs are cleaned up automatically
```

# HPA-Plasma {data-navmenu="Outputs" data-icon="fa-chart-bar"}
## Column {data-width=650}
### Scatter Plot
```{r}
plotlyOutput("hpa_plot", height = "550px")
```

## Column {data-width=350}
### Summary
```{r}
verbatimTextOutput("hpa_summary")
tableOutput("hpa_table")
```
```{r server_hpa, context="server"}

# ── Load & prep HPA data (done once, outside reactives) ──────────────────────
hpa <- read.csv(file.path(plasma_dir, "01_input", "HPA_CRCvHEALTHY_plasma.csv")) %>%
  dplyr::rename(
    gene      = Protein,
    logFC_HPA = NPX.fold.change,
    padj_HPA  = p.adjusted
  ) %>%
  mutate(logFC_HPA = as.numeric(logFC_HPA))

# ── Reactive: join HPA with limma results ────────────────────────────────────
hpa_compared <- reactive({
  my_results <- results_list[["Plasma"]][["Baseline_SURVvCTL"]] %>%
    dplyr::select(gene, logFC, adj.P.Val) %>%
    dplyr::rename(logFC_limma = logFC, padj_limma = adj.P.Val)

  inner_join(hpa, my_results, by = "gene") %>%
    mutate(
      direction_HPA   = ifelse(logFC_HPA   > 0, "up", "down"),
      direction_limma = ifelse(logFC_limma > 0, "up", "down"),
      agreement       = ifelse(direction_HPA == direction_limma, "Concordant", "Discordant")
    )
})

# ── Plot ──────────────────────────────────────────────────────────────────────
output$hpa_plot <- renderPlotly({
  compared <- hpa_compared()
  
  p <- ggplot(compared, aes(x = logFC_HPA, y = logFC_limma, color = agreement,
                             text = paste0("", gene, "<br>",
                                          "HPA logFC: ",   round(logFC_HPA,   3), "<br>",
                                          "Limma logFC: ", round(logFC_limma, 3), "<br>",
                                          "HPA padj: ",    signif(padj_HPA,  3), "<br>",
                                          "Agreement: ",   agreement))) +
    geom_point(size = 3, alpha = 0.85) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
    scale_color_manual(values = c("Concordant" = "#2166AC", "Discordant" = "#D62728")) +
    labs(
      x     = "CRC v Healthy logFC (HPA)",
      y     = "SURV v CTL logFC (ours)",
      title = "HPA vs Our Results",
      color = "Direction"
    ) +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
  
  ggplotly(p, tooltip = "text")
})

# ── Summary text ──────────────────────────────────────────────────────────────
output$hpa_summary <- renderText({
  compared <- hpa_compared()
  paste0(
    "Proteins in HPA:        ", nrow(hpa), "\n",
    "Proteins in limma:      ", nrow(results_list[["Plasma"]][["Baseline_SURVvCTL"]]), "\n",
    "Overlap:                ", nrow(compared), "\n",
    "Concordant:             ", sum(compared$agreement == "Concordant"), "\n",
    "Discordant:             ", sum(compared$agreement == "Discordant")
  )
})

# ── Table: HPA proteins not in your data ─────────────────────────────────────
output$hpa_table <- renderTable({
  anti_join(hpa, results_list[["Plasma"]][["Baseline_SURVvCTL"]], by = "gene") %>%
    dplyr::select(gene, logFC_HPA, padj_HPA) %>%
    dplyr::rename("Gene" = gene, "HPA logFC" = logFC_HPA, "HPA padj" = padj_HPA) %>%
    mutate(across(where(is.numeric), ~signif(.x, 3)))
}, caption = "HPA proteins not in your dataset")
```

